Insecticide resistance evolution with mixtures and sequences : a model-based explanation.

### 2017-09-21 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
paper2 tweaking

Our modelling approach focuses on the change in frequency of a single resistance gene for each insecticide. We assume both that genes conferring resistance are already present in the population and that resistance to each insecticide is coded by a single gene. These assumptions are consistent with previous modelling work. In support of the first assumption insecticide resistance has been termed 'pre-adaptive' meaning that resistance alleles are present in very low numbers even prior to exposure to novel insecticides [@Ffrench-Constant2013][@Liu2015] as has been documented in blowflies [Hartley2006]. In support of using a single gene representation of resistance evidence suggests that whilst polygenic resistance is common, control failures, particularly those due to target site resistance, are mostly due to single major genes [@Ffrench-Constant2013].

@@

**Table 1. Parameters inflencing the development of insecticide resistance**

@@

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | -------- | ---------------------------------------------
1. Effectiveness | faster | reduced fitness of SS in presence of insecticide
2. Exposure | faster | reduced fitness of SS overall
3. Dominance of restoration | faster | increased fitness of SR in presence of insecticide
4. Resistance restoration | faster | increased fitness of RR in presence of insecticide
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of RR in absence of insecticide

@@

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | ---------- | ---------------------------------------------
1. Effectiveness | **slower** | one insecticide reduces the fitness of individuals resistant to the other thereby reducing selection pressure from the other
2. Exposure | faster (but less than for single) | reduced fitness of individuals susceptible to one insecticide increases selection pressure for that resistance. However at the same time selection pressure is reduced by reduced fitness of resistant individuals caused by the other insecticide
3. Dominance of restoration | faster | increased fitness of heterozygotes
4. Resistance restoration | faster | increased fitness of resistants
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of resistants in absence of insecticide

@@

When single insecticide use was investigated (Figs 3-4) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution resistance, whereas higher values for cost led to slower evolution of resistance (Fig 4B).

These mechanisms, explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to figure 1 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

#### Two insecticides
Responses to changing input parameters are different when two insecticides are used either in sequence or in mixtures as compared to when used alone. These differences can help us to understand the dynamics of resistance evolution. In each panel of Figures 5-9 the solid lines, indicating that an insecticide is being used at the same time as another in a mixture, are always shallower than the same colour dashed line indicating the insecticide is being used in isolation as part of a sequence. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Because resistant individuals are being killed, the selection pressure for that resistance is lower. Thus each insecticide reduces the selection pressure for resistance to the other and could be said to 'protect' the other. This protection means that evolution of resistance to one insecticide is always slower when it is used in a mixture than when used alone. This protection has been termed multiple intra-generational killing [@Consortium2013] because individuals have the potential to be killed by more than one insecticide.

@@

Fig 5B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal, resistance will be expected to increase faster to the more effective insecticide (the red curve in this case). The more effective insecticide prompts both a) higher selection pressure for its own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide. The more effective insecticide has a faster evolution of resistance than the less effective one but this is not caused by the presence of the less effective one. (Indeed the presence of the less effective one is still slowing the evolution of resistance to the more effective one, as indicated by the red solid curve for use in the mixture being shallower than the red dashed curve for use alone as part of the sequence).

@@

#### Costs of resistance
Fitness costs of resistance have been widely reported(review in [@Kliot2012]) but it has been suggested that few of these are directly relevant to control failure in the field [@ffrench-Constant2017]. Because of this likely rarity of costs we did not include them in most of our analyses (with the exception of Fig 9).

@@

The resistance curves for two insecticides in a mixture that differ only in their cost of resistance, diverge more as the frequency of resistance increases (Fig 9C). This makes intuitive sense, as the frequency of resistance genes increases the impact of fitness costs associated with those genes becomes greater. Thus differences due to cost are less noticeable at low resistance frequencies and become more noticeable at higher resistance frequencies. Thus if a lower resistance threshold was used cost would have less of an effect and if a higher resistance threshold was used cost would have more of an effect. Also note that if the costs are set even higher than we set them here (e.g. above 0.2) they can maintain resistance frequencies below 1 or lead to their decline. However, we do not expect such high costs to occur in operational settings.

Declines in resistance frequencies are likely to occur when there are costs of resistance and the insecticide is not being used [@ffrench-Constant2017]. Thus the frequency of resistance for the first insecticide in a sequence, after it has been replaced, may decline to a level where the first insecticide becomes operationally useful again. Consequently, a repeated sequential strategy could offer benefits over mixtures not considered here. To quantify these benefits one would need to use a more detailed metric than the time-to-resistance we have used. We quantified the number of generations until resistance thresholds are first reached. Other measures for quantifying resistance over time can be imagined, for example the mean resistance or proportion of generations below a threshold.

@@

A subsequent model [@Roush1998] lead to the following more pessimistic conclusion about the potential of mixtures : “As a result of incomplete coverage and residue decay, the mortality of susceptible homozygotes is rarely consistently high enough for pesticide mixtures to be effective” [@Roush1998]. This "mortality of susceptible homozygotes" is equivalent to our insecticide effectiveness. Again this points to the importance of effectiveness and the question of whether the effectiveness levels needed to make mixtures better are attainable operationally. Our result showing that decreasing exposure favours mixtures over sequences is also consistent with figure 3 from that paper [@Roush1998].

One aspect of the results presented here can seem to contradict published advice and has been questioned when we have presented model results. Our model predicts that lower insecticide effectiveness or exposure will lead to slower spread of resistance in sole use and sequences (Figs 3 & 5, Table 2). This is consistent with expectations that lower effectiveness and exposure would lead to lower mortality of susceptibles and thus lower selection pressure for resistance [e.g. @Roush1998]. This can seem to contradict recommendations that poor implementation of insecticide interventions can promote the development of resistance. e.g. from [@FAO2012] "Certain pest control practices have consistently been shown to exacerbate the loss of susceptible pest populations and the development of resistance. These include: ... - the use of application rates that are below or above those recommended on the label, - poor coverage of the area being treated ...".

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? The explanation for this apparent difference is in the effect of dominance. Poor application can reduce the mortality of heterozygotes (RS) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [Bourguet2000][@Barbosa2011] [@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research into how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer.

@@

#### Caveats
There is much evidence of the involvement of multiple genes in resistance mechanisms. Our model, in common with most previous ones, assumes a single ‘resistance’ gene per insecticide. In most cases we believe this is reasonable. Target-site resistance has been shown to be mostly determined by single genes, and control failures in general have mostly been attributed to the effect of single major genes [@Ffrench-Constant2004] [@Ffrench-Constant2013]. Control doses of insecticides in the field are likely to lead to the selection of single major genes whereas the lower doses used to select laboratory strains are more likely to lead to polygenic resistance [@Ffrench-Constant2013]. To represent polygenic resistance a different 'quantitative genetic' approach would be required. A quantitative genetic model would likely lead to a more rapid initial evolution of resistance, a slower middle phase and a continuing increase. If polygenic resistance is shown in future to be more of a problem then these results could be compared to those of quantitative genetic models.

>>>>

Our modelling approach focuses on the change in frequency of a single resistance gene for each insecticide. We assume both that genes conferring resistance are already present in the population and that resistance to each insecticide is coded by a single gene. These assumptions are consistent with previous modelling work. In support of the first assumption insecticide resistance has been termed 'pre-adaptive' meaning that resistance alleles are present in very low numbers even prior to exposure to novel insecticides [@Ffrench-Constant2013][@Liu2015] as has been documented in blowflies [@Hartley2006]. In support of using a single gene representation of resistance evidence suggests that whilst polygenic resistance is common, control failures, particularly those due to target site resistance, are mostly due to single major genes [@Ffrench-Constant2013].

@@

**Table 1. Parameters influencing the development of insecticide resistance**

@@

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | ------ | ---------------------------------------------
1. Effectiveness | faster | reduced fitness of SS & SR exposed to insecticide
2. Exposure | faster | reduced fitness of SS & SR overall
3. Dominance of restoration | faster | increased fitness of SR exposed to insecticide
4. Resistance restoration | faster | increased fitness of RR & SR exposed to insecticide
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of RR & SR not exposed to insecticide

@@

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | ---------- | ---------------------------------------------
1. Effectiveness | **slower** | one insecticide reduces the fitness of individuals resistant to the other thereby reducing selection pressure for the other
2. Exposure | faster (but less than for insecticide used alone) | reduced fitness of individuals susceptible to one insecticide increases selection pressure for that resistance. However at the same time selection pressure is reduced by reduced fitness of resistant individuals caused by the other insecticide
3. Dominance of restoration | faster | increased fitness of heterozygotes
4. Resistance restoration | faster | increased fitness of resistants
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of resistants in absence of insecticide

@@

When single insecticide use was investigated (Figs 3-4) higher values of parameters that increased the selective advantage of resistance all led to more rapid evolution resistance, whereas higher values for cost led to slower evolution of resistance (Fig 4B). Mechanisms explaining the evolution of resistance to single insecticide use, are summarised in Table 2. Referring to figure 1 can help explain these responses. Exposure and effectiveness act by decreasing the fitness of SS and SR more than RR, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR among those mosquitoes  not exposed.

#### Two insecticides
Responses to input parameters are different when two insecticides are used either in sequence or in mixtures as compared to when used alone. These differences can help us to understand the dynamics of resistance evolution. In each panel of Figures 5-9 the solid lines, indicating that an insecticide is being used at the same time as another in a mixture, are always shallower than the same colour dashed line indicating the insecticide is being used in isolation as part of a sequence. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Because resistant individuals are being killed, the selection pressure for that resistance is lower. Thus each insecticide reduces the selection pressure for resistance to the other and could be said to 'protect' the other. This protection means that evolution of resistance to one insecticide is always slower when it is used in a mixture than when used alone. This protection has been termed multiple intra-generational killing [@Consortium2013] because individuals have the potential to be killed by more than one insecticide.

@@

Fig 5B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal, resistance will be expected to increase faster to the more effective insecticide (the red curve in this case). The more effective insecticide prompts both a) higher selection pressure for its own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide. The more effective insecticide has a faster evolution of resistance than the less effective one but this is not caused by the presence of the less effective one. Indeed the presence of the less effective one is still slowing the evolution of resistance to the more effective one, as indicated by the red solid curve for use in the mixture being shallower than the red dashed curve for use alone as part of the sequence.

@@

#### Costs of resistance
Fitness costs of resistance have been widely reported(review in [@Kliot2012]) but it has been suggested that few of these are directly relevant to control failure in the field [@Ffrench-Constant2017]. Because of this likely rarity of costs we did not include them in most of our analyses (with the exception of Fig 9).

@@

The resistance curves for two insecticides in a mixture that differ only in their cost of resistance, diverge more as the frequency of resistance increases (Fig 9C). This makes intuitive sense, as the frequency of resistance genes increases the impact of fitness costs associated with those genes also increases. Thus differences due to cost are less noticeable at low resistance frequencies and become more noticeable at higher resistance frequencies. Thus if a lower resistance threshold was used cost would have less of an effect and if a higher resistance threshold was used cost would have more of an effect. Also note that if the costs are set even higher than we set them here (e.g. above 0.2) they can maintain resistance frequencies below 1 or lead to their decline. However, we do not expect such high costs to occur in operational settings.

Declines in resistance frequencies are likely to occur when there are costs of resistance and the insecticide is not being used [@Ffrench-Constant2017]. Thus the frequency of resistance for the first insecticide in a sequence, after it has been replaced, may decline to a level where the first insecticide becomes operationally useful again. Consequently, a repeated sequential strategy could offer benefits over mixtures not considered here. To quantify these benefits one would need to use a more detailed metric than the time-to-resistance we have used. We quantified the number of generations until resistance thresholds are first reached. Other measures for quantifying resistance over time can be imagined, for example the mean resistance or proportion of generations below a resistance threshold.

@@

A subsequent model lead to the following more pessimistic conclusion about the potential of mixtures : “As a result of incomplete coverage and residue decay, the mortality of susceptible homozygotes is rarely consistently high enough for pesticide mixtures to be effective” [@Roush1998]. This "mortality of susceptible homozygotes" is equivalent to our insecticide effectiveness. Again this points to the importance of effectiveness and the question of whether the effectiveness levels needed to make mixtures better are attainable operationally. Consistent with our results [@Roush1998] also showed (Figure 3) that decreasing exposure favours mixtures over sequences.

#### Poor implementation
One aspect of the results presented here can seem to contradict published advice and has been questioned when we have presented model results. Our model predicts that lower insecticide effectiveness or exposure will lead to slower spread of resistance in sole use and sequences (Figs 3 & 5, Table 2). This is consistent with expectations that lower effectiveness and exposure would lead to lower mortality of susceptibles and thus lower selection pressure for resistance [e.g. @Roush1998]. This can seem to contradict recommendations that poor implementation of insecticide interventions can promote the development of resistance. e.g. from [@FAO2012] "Certain pest control practices have consistently been shown to exacerbate the loss of susceptible pest populations and the development of resistance. These include: ... - the use of application rates that are below or above those recommended on the label, - poor coverage of the area being treated ...".

Poor implementation would likely cause lower effectiveness of, and exposure to, insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? The explanation for this apparent difference is in the effect of dominance. Poor application can reduce the mortality of heterozygotes (SR) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Bourguet2000][@Barbosa2011][@FAO2012] see Fig 1 of [@Levick2017]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that higher effectiveness favours mixtures, higher exposure favours sequences and higher dominance favours neither (Table 4). More research into how the mortality of different genotypes (susceptible, resistant and heterozygous) is affected by poor application rates or insecticide degradation over time would help make this clearer.

@@

#### Caveats
There is much evidence of the involvement of multiple genes in resistance mechanisms. Our model, in common with most previous ones, assumes a single ‘resistance’ gene per insecticide. In most cases we believe this is reasonable. Target-site resistance has been shown to be mostly determined by single genes, and control failures in general have mostly been attributed to the effect of single major genes [@Roush1989][@Ffrench-Constant2004][@Ffrench-Constant2013]. Control doses of insecticides in the field are likely to lead to the selection of single major genes whereas the lower doses used to select laboratory strains are more likely to lead to polygenic resistance [@Ffrench-Constant2013]. To represent polygenic resistance a different 'quantitative genetic' approach would be required [@Via1986][@Firko1990]. We are only aware of preliminary published work applying a quantitative genetic approach to assessing insecticide use strategies in the context of polygenic resistance [@Via1986][@Gardner1998]. One suggestion is that evolution of resistance to one insecticide in a mixture could be associated with a decrease in tolerance to the other [@Via1986]. If it becomes clearer that polygenic resistance is an issue then we would welcome a comparison of the results presented here to those of quantitative genetic models.
